---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dev = "ragg_png",
  dpi = 300
)
```

# econgoods

<!-- badges: start -->
<!-- badges: end -->

econgoods provides utility functions and composite goods in economics.

## Installation

You can install the development version of econgoods from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("UchidaMizuki/econgoods")
```

## Examples

```{r load-libraries, message=FALSE, warning=FALSE}
library(econgoods)
library(tidyverse)
```

### Calibrate utility functions

- Utility functions:
  - `util_cobb_douglas()`: Cobb-Douglas utility function
  - `util_leontief()`: Leontief utility function
  - `util_ces()`: Constant elasticity of substitution (CES) utility function
- `util_calibrate()`: Fit parameters of utility functions to the given prices 
and quantities.

```{r calibrate-utility-functions-sample-data}
# Sample data
prices <- c(2, 1)
income <- 6

x <- 2
y <- util_2goods_budget(prices, income)(x)
quantities <- c(x, y)
```

```{r calibrate-utility-functions}
cobb_douglas <- util_cobb_douglas() |> 
  util_calibrate(prices, quantities)
cobb_douglas
leontief <- util_leontief() |> 
  util_calibrate(prices, quantities)
leontief
ces_minus_1_5 <- util_ces(substitution = -1.5) |> 
  util_calibrate(prices, quantities)
ces_minus_1_5
```

### Indifference curve and budget line for two goods

- `util_2goods_indifference()` returns the function of indifference curve.
- `util_2goods_budget()` returns the function of budget line.

```{r}
util_2goods_indifference(cobb_douglas, quantities)(1:6)
util_2goods_budget(prices, income)(1:6)
```
#### Advanced plotting examples

```{r plot-indifference-curve-and-budget-line, echo=FALSE}
# You need to install dev version of geomtextpath
# - https://github.com/AllanCameron/geomtextpath/issues/114
library(geomtextpath)

theme_set(theme_bw())

limits <- c(0, max(income / prices))
stat_2goods_indifference <- function(f, ...) {
  stat_function(geom = "textpath",
                fun = \(x) util_2goods_indifference(f, quantities)(x) |> 
                  replace_na(1e6),
                label = pillar::obj_sum(f) |> 
                  # https://github.com/AllanCameron/geomtextpath/issues/75
                  str_replace_all("-", "\u2212"),
                hjust = 1,
                xlim = limits,
                ...)
}
stat_2goods_budget <- function(...) {
  stat_function(geom = "textpath",
                fun = util_2goods_budget(prices, income),
                label = "Budget line",
                hjust = 0.2,
                xlim = c(0, income / prices[1]),
                ...)
}
geom_quantities <- function(...) {
  geom_point(aes(quantities[[1]], quantities[[2]]),
             ...)
}
ggplot() +
  geom_hline(yintercept = 0,
             color = "dimgray") +
  geom_vline(xintercept = 0,
             color = "dimgray") +
  stat_2goods_indifference(cobb_douglas,
                           color = "red4") +
  stat_2goods_indifference(leontief,
                           n = 201,
                           color = "red4") +
  stat_2goods_indifference(ces_minus_1_5,
                           color = "red4") +
  stat_2goods_budget(color = "blue4") +
  geom_quantities(color = "blue4",
                  size = 2) +
  scale_x_continuous("Quantity of good X") +
  scale_y_continuous("Quantity of good Y") +
  tune::coord_obs_pred(xlim = limits,
                       ylim = limits)
```

```{r, plot-utility-level, warning=FALSE, fig.width=7, fig.height=6, echo=FALSE}
ces_0_75 <- util_ces(substitution = 0.75) |> 
  util_calibrate(prices, quantities)
get_data_2goods_utility <- function(f) {
  f |> 
    purrr::map(
      \(f) {
        expand_grid(x = seq(0, limits[[2]], length.out = 100),
                    y = seq(0, limits[[2]], length.out = 100)) |>
          rowwise() |> 
          mutate(z = f(c(x, y))) |> 
          ungroup()
      }
    ) |> 
    set_names(map_chr(f, pillar::obj_sum)) |> 
    bind_rows(.id = "utility") |> 
    mutate(across(utility, as_factor))
}
get_data_2goods_utility(c(leontief, ces_minus_1_5, cobb_douglas, ces_0_75)) |> 
  ggplot(aes(x, y)) +
  geom_contour_filled(aes(z = z)) +
  stat_2goods_budget(color = "white") +
  geom_quantities(color = "white") +
  scale_x_continuous("Quantity of good X") +
  scale_y_continuous("Quantity of good Y") +
  facet_wrap(~ utility,
             ncol = 2) +
  tune::coord_obs_pred()
```
