---
  output: github_document
---
  
<!-- README.md is generated from README.Rmd. Please edit that file -->
  
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dev = "ragg_png",
  dpi = 300
)
```

# econgoods

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->
  
econgoods provides utility functions and composite goods in economics.

## Installation

You can install the development version of econgoods from [GitHub](https://github.com/) with:
  
``` r
# install.packages("devtools")
devtools::install_github("UchidaMizuki/econgoods")
```

## Examples for utility functions

```{r load-libraries, message=FALSE, warning=FALSE}
library(econgoods)
library(tidyverse)
```

### Calibrate utility functions

- Utility functions:
  - `util_cobb_douglas()`: Cobb-Douglas utility function
  - `util_leontief()`: Leontief utility function
  - `util_ces()`: Constant elasticity of substitution (CES) utility function
- `util_calibrate()`: Fit parameters of utility functions to the given prices 
and quantities.

```{r calibrate-utility-functions-sample-data}
# Sample data
prices <- c(4, 2)
income <- 12

quantity_x <- 2
quantity_y <- util_2goods_budget(prices, income)(quantity_x)
quantities <- c(quantity_x, quantity_y)
```

```{r calibrate-utility-functions}
cobb_douglas <- util_cobb_douglas() |> 
  util_calibrate(prices, quantities)
cobb_douglas
leontief <- util_leontief() |> 
  util_calibrate(prices, quantities)
leontief
ces_minus_1_5 <- util_ces(substitution = -1.5) |> 
  util_calibrate(prices, quantities)
ces_minus_1_5
```

### Indifference curve and budget line for two goods

- `util_2goods_indifference()` returns the function of indifference curve.
- `util_2goods_budget()` returns the function of budget line.

```{r indifference-curve-and-budget-line}
util_2goods_indifference(cobb_douglas, quantities)(1:6)
util_2goods_budget(prices, income)(1:6)
```

#### Sample plots

```{r plot-indifference-curve-and-budget-line, echo=FALSE}
# You need to install dev version of geomtextpath
# - https://github.com/AllanCameron/geomtextpath/issues/114
library(geomtextpath)

theme_set(theme_bw())

limits <- c(0, max(income / prices))
stat_indifference <- function(f, quantities,
                              label = pillar::obj_sum(f) |> 
                                # https://github.com/AllanCameron/geomtextpath/issues/75
                                str_replace_all("-", "\u2212"),
                              ...) {
  fun <- util_2goods_indifference(f, quantities, otherwise = 1e6)
  if (is.null(label)) {
    stat_function(fun = fun,
                  xlim = limits,
                  ...)
  } else {
    stat_function(geom = "textpath",
                  fun = fun,
                  label = label,
                  hjust = 1,
                  xlim = limits,
                  ...) 
  }
}
stat_2goods_budget <- function(prices, income, 
                               label = "Budget line",
                               ...) {
  xlim <- c(0, income / prices[1])
  fun <- util_2goods_budget(prices, income)
  if (is.null(label)) {
    stat_function(fun = fun,
                  xlim = xlim,
                  ...)
  } else {
    stat_function(geom = "textpath",
                  fun = fun,
                  label = label,
                  hjust = 0.2,
                  xlim = xlim,
                  ...)
  }
}
geom_point_quantities <- function(quantities, ...) {
  geom_point(aes(quantities[[1]], quantities[[2]]),
             ...)
}
ggplot() +
  geom_hline(yintercept = 0,
             color = "dimgray") +
  geom_vline(xintercept = 0,
             color = "dimgray") +
  stat_indifference(cobb_douglas, quantities,
                    color = "red4") +
  stat_indifference(leontief, quantities,
                    n = 201,
                    color = "red4") +
  stat_indifference(ces_minus_1_5, quantities,
                    color = "red4") +
  stat_2goods_budget(prices, income,
                     color = "blue4") +
  geom_point_quantities(quantities,
                        color = "blue4",
                        size = 2) +
  scale_x_continuous("Quantity of good X") +
  scale_y_continuous("Quantity of good Y") +
  tune::coord_obs_pred(xlim = limits,
                       ylim = limits)
```

```{r, plot-utility-level, warning=FALSE, fig.width=7, fig.height=6, echo=FALSE}
ces_0_75 <- util_ces(substitution = 0.75) |> 
  util_calibrate(prices, quantities)
get_data_2goods_utility <- function(f) {
  f |> 
    purrr::map(
      \(f) {
        expand_grid(quantity_x = seq(0, limits[[2]], length.out = 100),
                    quantity_y = seq(0, limits[[2]], length.out = 100)) |>
          rowwise() |> 
          mutate(utility = f(c(quantity_x, quantity_y))) |> 
          ungroup()
      }
    ) |> 
    set_names(map_chr(f, pillar::obj_sum)) |> 
    bind_rows(.id = "utility_type") |> 
    mutate(across(utility_type, as_factor))
}
get_data_2goods_utility(c(leontief, ces_minus_1_5, cobb_douglas, ces_0_75)) |> 
  ggplot(aes(quantity_x, quantity_y)) +
  geom_contour_filled(aes(z = utility),
                      binwidth = 4) +
  stat_2goods_budget(prices, income,
                     color = "white") +
  geom_point_quantities(quantities,
                        color = "white") +
  scale_x_continuous("Quantity of good X") +
  scale_y_continuous("Quantity of good Y") +
  scale_fill_viridis_d("Utility") +
  facet_wrap(~ utility_type,
             ncol = 2) +
  tune::coord_obs_pred()
```

### Marginal utility for two goods

- `util_2goods_utility()` returns the function of total utility or marginal utility with a given quantity of good Y.
  - `gradient = FALSE`: Total utility (default)
  - `gradient = TRUE`: Marginal utility for good X

```{r marginal-utility}
# Total utility
util_2goods_utility(cobb_douglas, quantities[[2]])(1:6)
# Marginal utility
util_2goods_utility(cobb_douglas, quantities[[2]], gradient = TRUE)(1:6)
```
#### Sample plots

```{r plot-marginal-utility, echo=FALSE, fig.width=7, fig.height=4, warning=FALSE}
library(patchwork)

stat_utility <- function(f, 
                         gradient = FALSE,
                         label = pillar::obj_sum(f) |> 
                           # https://github.com/AllanCameron/geomtextpath/issues/75
                           str_replace_all("-", "\u2212"),
                         ...) {
  if (gradient) {
    fun <- util_2goods_utility(f, quantities[[2]], gradient = TRUE)
  } else {
    fun <- util_2goods_utility(f, quantities[[2]])
  }
  if (is.null(label)) {
    stat_function(fun = fun,
                  xlim = limits,
                  ...)
  } else {
    stat_function(geom = "textpath",
                  fun = fun,
                  label = label,
                  hjust = 1,
                  xlim = limits,
                  ...) 
  }
}

color <- RColorBrewer::brewer.pal(4, "Dark2")
plot_total_utility <- ggplot() +
  stat_utility(leontief,
               color = color[[1]]) +
  stat_utility(ces_minus_1_5,
               color = color[[2]]) +
  stat_utility(cobb_douglas,
               color = color[[3]]) +
  scale_x_continuous("Quantity of good X",
                     limits = c(1e-3, NA)) +
  scale_y_continuous("Total utility")
plot_marginal_utility <- ggplot() +
  stat_utility(leontief,
               gradient = TRUE,
               n = 201,
               color = color[[1]]) +
  stat_utility(ces_minus_1_5,
               gradient = TRUE,
               color = color[[2]]) +
  stat_utility(cobb_douglas,
               gradient = TRUE,
               color = color[[3]]) +
  scale_x_continuous("Quantity of good X",
                     limits = c(1e-3, NA)) +
  scale_y_continuous("Marginal utility of good X")

plot_total_utility + plot_marginal_utility
```

### Price effect for two goods

```{r price-effect}
prices_new <- c(2, 2)

# Price effect
quantities_new <- util_demand(cobb_douglas, prices_new,
                              income = income)
quantities_new
# Substitution effect
quantities_substitution <- util_demand(cobb_douglas, prices_new,
                                       utility = cobb_douglas(quantities))
quantities_substitution
# Income effect
quantities_new - quantities_substitution
```

#### Sample plots

```{r plot-price-effect, echo=FALSE}
geom_text_quantities <- function(quantities,
                                 label = str_c("(", 
                                               scales::label_number(accuracy = 1e-2)(quantities[[1]]), 
                                               ", ", 
                                               scales::label_number(accuracy = 1e-2)(quantities[[2]]), 
                                               ")"),
                                 hjust = -0.2,
                                 ...) {
  geom_text(aes(quantities[[1]], quantities[[2]]),
            label = label,
            hjust = hjust,
            ...)
}

ggplot() +
  geom_hline(yintercept = 0,
             color = "dimgray") +
  geom_vline(xintercept = 0,
             color = "dimgray") +
  stat_indifference(cobb_douglas, quantities,
                    color = "red4") +
  stat_indifference(cobb_douglas, quantities_new,
                    label = NULL,
                    color = "red4") +
  stat_2goods_budget(prices, income,
                     color = "blue4") +
  stat_2goods_budget(prices_new, income,
                     label = NULL,
                     color = "blue4") +
  stat_2goods_budget(prices_new, sum(prices_new * quantities_substitution),
                     label = NULL,
                     linetype = "dashed",
                     color = "blue4") +
  geom_point_quantities(quantities,
                        color = "blue4") +
  geom_point_quantities(quantities_new,
                        color = "blue4") +
  geom_point_quantities(quantities_substitution,
                        color = "blue4") +
  geom_text_quantities(quantities,
                       color = "blue4") +
  geom_text_quantities(quantities_new,
                       color = "blue4") +
  geom_text_quantities(quantities_substitution,
                       color = "blue4") +
  scale_x_continuous("Quantity of good X") +
  scale_y_continuous("Quantity of good Y") +
  tune::coord_obs_pred(xlim = limits,
                       ylim = limits)
```

## Examples for composite goods

- TODO
